---
title: "PML Project"
output: html_notebook
---

```{r, include=FALSE}
library(tidyverse)
library(caret)
library(glmnet)
```

## Links to Github
[Repository](https://github.com/rhartvik/bicep-curl-form)

[Compiled HTML file](https://rhartvik.github.io/bicep-curl-form/)

# Classifying exercise forms

## Creating rest, training, validation sets
```{r}
labeled_data <- read.csv("pml-training.csv")

inBuild <- createDataPartition(labeled_data$classe, p=0.7, list=FALSE)
validation <- labeled_data[-inBuild,]
buildData <- labeled_data[inBuild,]

inTrain <- createDataPartition(buildData$classe, p=0.7, list=FALSE)
training <- buildData[-inTrain,]
testing <- buildData[inTrain,]
```
```{r, include=FALSE}
dim(training)
dim(testing)
dim(validation)
```
## How I built my model

The dataset started with `r dim(training)[2]` columns.
In order to produce a model that will train in a reasonable amount of time, I want to select only a subset of the columns. I will focus on eliminating columns that are unlikely to provide value to the model.

1. Remove columns with little variance results:
```{r}
training1 <- training[,-nearZeroVar(training)]
```
 `r dim(training1)[2]` columns remain.


2. Remove columns that are null for too many rows:
```{r}
percentage_provided_per_classe <- training1 %>% group_by(classe) %>% summarise_each(funs(sum(!is.na(.))/n()))
percentage_provided <- training1 %>% summarise_each(funs(sum(!is.na(.))/n()))
incomplete_columns <- colnames(percentage_provided[,(percentage_provided[1,]) < 1])
training2 <- training1 %>% select(-incomplete_columns)
```
 `r dim(training2)[2]` complete columns remain.



3. Investigate for highly-correlated data
```{r}
numeric_cols <- unlist(lapply(training2, is.numeric))
numeric_data <- training2[ , numeric_cols]
numeric_data <- numeric_data[,-1] # Remove the row number

cors <- abs(cor(numeric_data))
diag(cors) <- 0
which (cors > 0.95, arr.ind = T)
```
I'll remove one of each pair of high-related columns, since the second in the pair is very unlikely to add additional information that wasn't captured in the first.
```{r}
coorelated_cols <- colnames(numeric_data %>% select(7,4,5))
training3 <- training2 %>% select(-coorelated_cols)
```
We still have `r dim(training3)[2]` columns.

Remove more correlated columns
```{r}
cors <- abs(cor(numeric_data %>% select(-coorelated_cols)))
diag(cors) <- 0
which (cors > 0.80, arr.ind = T)
```

```{r}
coorelated_cols2 <- colnames(numeric_data %>% select(8,9,18,24,25,28,29))
training4 <- training3 %>% select(-coorelated_cols2)
```
We still have `r dim(training4)[2]` columns.

4. Remove columns that don't contain real data
The formated date is redundant with the other datetimestamps and the row number is administrative.
```{r}
training5 <- training4 %>% select(-X, -cvtd_timestamp)
```


Normalize the training set
```{r}
preProcess <- preProcess(training5, method=c("center", "scale"))
training5_norm <- predict(preProcess,training5)
```


Run a couple different training methods
```{r}
model1 <- train(training5_norm %>% select(-classe), training$classe, method="treebag") # (Training) Accuracy: 0.9849203  Kappa: 0.9809043

model2 <- train(training5_norm %>% select(-classe, -user_name, -num_window), training$classe, method="treebag")

```

## Cross Validation
Prep the test set:
```{r}
testing5 <- testing %>% select(colnames(training5))
testing5_norm <- predict(preProcess, testing5)
```


```{r}
model1_test_predictions <- predict(model1,testing5_norm)
(model1_test_cmatrix <- confusionMatrix(model1_test_predictions,testing5_norm$classe))
# Accuracy : 0.9921      
model2_test_predictions <- predict(model2,testing5_norm)
(model2_test_cmatrix <- confusionMatrix(model2_test_predictions,testing5_norm$classe))

model1_test_cmatrix[3]
model2_test_cmatrix[3]
```
Model 1 did better than model 2.

Test with another random sample:
```{r}
inTrain <- createDataPartition(buildData$classe, p=0.7, list=FALSE)
training_v2 <- buildData[-inTrain,]
testing_v2 <- buildData[inTrain,]

testing_v2_5 <- testing_v2 %>% select(colnames(training5))
testing_v2_5_norm <- predict(preProcess, testing_v2_5)

training_v2_5 <- training_v2 %>% select(colnames(training5))
training_v2_norm <- predict(preProcess, training_v2_5)


model1_v2 <- train(testing_v2_5_norm %>% select(-classe), testing_v2_5_norm$classe, method="treebag")
model2_v2 <- train(testing_v2_5_norm %>% select(-classe, -user_name, -num_window), testing_v2_5_norm$classe, method="treebag")

model1_test_v2_predictions <- predict(model1_v2,testing5_norm)
(model1_test_v2_cmatrix <- confusionMatrix(model1_test_v2_predictions,testing5_norm$classe))
  
model2_test_v2_predictions <- predict(model2_v2,testing5_norm)
(model2_test_v2_cmatrix <- confusionMatrix(model2_test_v2_predictions,testing5_norm$classe))
```
Using a new random sample, model 1 still had a better in sample error rate than model 2, with an accuracy of around 99.8%.


## Expected out of sample error
The in sample error was low: Accuracy of 98% one models.
To predict the out of sample error, I will test the models on the validation set (which was not used to develop the models).
```{r}
validation_5 <- validation %>% select(colnames(training5))
validation_5_norm <- predict(preProcess, validation_5)

model1_validation_v2_predictions <- predict(model1_v2,validation_5_norm)
(model1_validation_v2_cmatrix <- confusionMatrix(model1_validation_v2_predictions,validation_5_norm$classe))
  
model2_validation_v2_predictions <- predict(model2_v2,validation_5_norm)
(model2_test_v2_cmatrix <- confusionMatrix(model2_validation_v2_predictions,validation_5_norm$classe))
```
Out of sample accuracy is estimated at 99.5%, which is worse than the in sample accuracy, as expected, since in sample error rates should be less than out of sample error rates.


## Predict 20 different test cases
```{r}
final_test_set <- read.csv("pml-testing.csv")
final_test_set6 <- final_test_set %>% select(colnames(training5 %>% select(-classe)))
final_test_set6_norm <- predict(preProcess, final_test_set6)

model1_test_predictions <- predict(model1,final_test_set6_norm)
(model1_test_cmatrix <- confusionMatrix(model1_test_predictions,testing6_norm$classe))
```
